
\documentclass{article}

\usepackage[preprint]{neurips_2026}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{array}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\title{Algorithms as Neural Modules:\\
  Exact Mathematical Structure in Differentiable Networks}

\author{%
  Stuart Whipp\\
  Independent Research\\
  \texttt{swhipp87@gmail.com}
}

\begin{document}

\maketitle

\begin{abstract}
We present a family of zero-parameter neural modules that encode classical mathematical
algorithms---Pascal's triangle, the Sieve of Eratosthenes, and binomial expansion---as
fixed-weight networks composable with learned components.
Unlike supervised models, these modules solve their target tasks with provable correctness
and require no training data.
We show that tasks which defeat standard MLPs entirely---primality detection, exact
polynomial coefficient recovery---are trivially solved by the corresponding
zero-parameter module.
We further demonstrate a mixture-of-experts (MoE) model in which a learned router
discovers, without supervision, that it should delegate arithmetic queries to the
structured expert: the router weight toward the structured sieve rises from 0.58 to
0.94 over training, while the blind MLP expert fails to extrapolate beyond its training
range.
The structured expert, by contrast, is exact across all inputs by construction.
A third expert slot, reserved for differentiable forward chaining over Horn-clause
knowledge bases, is already implemented in the accompanying library
(\texttt{polyweave.reasoning}) and constitutes a natural extension of this framework
to logical inference.
All modules are released as \texttt{polyweave.maths}
(\url{https://github.com/swhipp/autodidact-hypernet}).
\end{abstract}

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}
% ─────────────────────────────────────────────────────────────────────────────

The Peano axioms establish the natural numbers from two primitives: zero and a successor
function. Every arithmetic fact about $\mathbb{N}$ follows by induction from these
primitives alone. Our observation, and the starting point of this paper, is that the
same incremental construction is available inside a neural network. A single frozen
value and a fixed local rule---encoded as a weight matrix---suffice to generate Pascal's
triangle, Bernoulli's triangle, and the binomial coefficients of any $(Ax + By)^n$,
exactly and without training.

This matters for a practical reason. There is growing interest in equipping language
models with the ability to call external tools---calculators, symbolic solvers, code
interpreters---for tasks that require exact computation \citep{toolformer2023,
react2022}. We argue that a complementary capability is equally important: the ability
to perform a \emph{finite, well-defined class} of arithmetic operations \emph{internally},
before incurring the latency and brittleness of an external call. This is analogous
to human mental arithmetic---a person does not reach for a calculator to square a small
integer---and we suggest that structured neural modules provide a substrate for this kind
of reliable internal computation in artificial systems.

Our modules are implemented as parameter-free \texttt{nn.Module} subclasses in PyTorch.
They compose naturally with learned components and are differentiable with respect to
their continuous hyperparameters, enabling joint optimisation in larger systems.
We contribute:

\begin{enumerate}
  \item Three zero-parameter modules---\texttt{PascalTriangle},
    \texttt{BinomialExpansion}, and \texttt{DifferentiableSieve}---that solve their
    tasks with provable 100\% accuracy and no training data (Section~\ref{sec:modules}).
  \item A \texttt{BernoulliTriangle} module that encodes simplex and cake-number
    sequences as a column-wise cumulative sum of Pascal, further illustrating the
    combinatorial richness accessible from a single fixed-weight primitive
    (Section~\ref{sec:modules}).
  \item A mixture-of-experts framework in which a learned router empirically discovers
    that it should trust the structured sieve expert, with the router weight rising
    monotonically from initialisation and settling near one (Section~\ref{sec:tier2}).
  \item An honest comparison: an MLP baseline trained on the same tasks, showing the
    qualitative failure modes that structured modules avoid (Section~\ref{sec:tier1}).
  \item Release of all modules as \texttt{polyweave.maths}, with the existing
    \texttt{polyweave.reasoning} forward chainer constituting a ready-made logic expert
    for future MoE extension (Section~\ref{sec:discussion}).
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Related Work}
\label{sec:related}
% ─────────────────────────────────────────────────────────────────────────────

\paragraph{Neural arithmetic.}
Neural Arithmetic Logic Units \citep{nalu2018} learn to perform arithmetic operations
through weight-constrained modules. Neural GPU \citep{neuralgpu2015} solves
multiplication via convolutional recurrence. Our approach is complementary: rather than
learning to approximate arithmetic, we encode exact algorithms as frozen network
structure.

\paragraph{Differentiable programming and neurosymbolic reasoning.}
DeepProbLog \citep{deepproblog2018}, Logical Neural Networks \citep{lnn2020}, and
Neural Theorem Provers \citep{ntp2017} embed symbolic reasoning in differentiable
systems. \citet{differentiableILP2017} learn inductive logic programs end-to-end. We
focus on the numerical and combinatorial side of this space rather than logical
inference, though our accompanying \texttt{polyweave.reasoning} module bridges toward
the latter.

\paragraph{Mixture of experts.}
Standard MoE architectures \citep{moe2017,switch2022} route between learned experts.
We route between a \emph{frozen structured expert} and a learned expert---an asymmetric
setting in which the router must discover that exact structure exists, not merely which
of two approximators performs better.

\paragraph{Tool use and internal computation.}
Toolformer \citep{toolformer2023} and ReAct \citep{react2022} teach language models to
invoke external tools for arithmetic and search. Our work proposes an orthogonal
direction: certain computations should be handled \emph{internally} by structured
modules, reserving external calls for problems outside their scope.

\paragraph{This paper in context.}
This is the third in a series. Paper~1 \citep{paper1} established that multiplicative
(Sigma-Pi) neurons are recruited in proportion to the nonlinearity of the mapping they
are asked to learn, and introduced \texttt{pi\_scale} as a diagnostic. Paper~2
\citep{paper2} showed that transformer feed-forward blocks are largely linearly
recoverable and can be compressed into a single polynomial layer. The present paper
extends the \emph{structured-computation} theme to exact arithmetic modules.

% ─────────────────────────────────────────────────────────────────────────────
\section{Mathematical Modules}
\label{sec:modules}
% ─────────────────────────────────────────────────────────────────────────────

All modules are parameter-free \texttt{nn.Module} subclasses. Weights are registered
as frozen buffers; no gradient flows through them. Each module can be placed inside a
larger differentiable graph and is compatible with \texttt{torch.compile}.

\subsection{Pascal's Triangle (\texttt{PascalTriangle})}

Pascal's triangle satisfies $C(n,k) = C(n{-}1,k{-}1) + C(n{-}1,k)$.
We implement this as a fixed-weight 2-D convolution with kernel
$\mathbf{K} = \bigl[\begin{smallmatrix}1&1\\0&0\end{smallmatrix}\bigr]$,
applied iteratively with top-left zero-padding. Summing all intermediate states gives
the full triangle because intermediate $\mathbf{x}_k$ is non-zero only at row $k$:

\begin{equation}
  \mathbf{x}_{k+1}[i,j] = \mathbf{x}_k[i{-}1,j{-}1] + \mathbf{x}_k[i{-}1,j],
  \quad
  \text{output} = \sum_{k=0}^{N-1} \mathbf{x}_k.
\end{equation}

The output is exact (integer-valued floating-point) for $n \leq 15$.
Figure~\ref{fig:pascal} shows the result as a log-scale heatmap.

\subsection{Binomial Expansion (\texttt{BinomialExpansion})}

Given scalars $A$, $B$ and exponent $n$, the binomial theorem gives:
\begin{equation}
  (Ax + By)^n = \sum_{k=0}^{n} \binom{n}{k} A^{n-k} B^k\, x^{n-k} y^k.
\end{equation}
We compute this by composing \texttt{PascalTriangle} with frozen exponent lookup tables.
The module takes $(A, B, n)$ and returns the exact coefficient vector of length
\texttt{num\_rows}, with trailing zeros beyond position $n$.
Zero learnable parameters; output matches symbolic computation to floating-point
precision on the supported integer domain.

\subsection{Bernoulli's Triangle (\texttt{BernoulliTriangle})}

Bernoulli's triangle contains the partial row sums of Pascal:
$B(n,k) = \sum_{j=0}^{k} C(n,j)$.
Notable sequences emerge along columns: naturals ($k{=}1$), 2-D lazy-caterer numbers
($k{=}2$: $1,2,4,7,11,\ldots$), 3-D cake numbers ($k{=}3$: $1,2,4,8,15,26,\ldots$),
and $k$-simplex numbers in general.
We implement $B$ as \texttt{torch.cumsum} applied column-wise to the Pascal output---a
natural fixed-weight operation consistent with the frozen-buffer pattern throughout.

\subsection{Differentiable Sieve (\texttt{DifferentiableSieve})}
\label{sec:sieve}

The Sieve of Eratosthenes marks composites by sweeping multiples of each prime
$p \leq \lfloor\sqrt{N}\rfloor$.
We implement each prime's sweep as a frozen \emph{comb buffer}---a tensor of length
$N{+}1$ with 1s at positions $0, p, 2p, \ldots$ (position $p$ cleared, since $p$ is
prime). Composite scores are combined via the probabilistic OR formula:
\begin{equation}
  \text{composite}(n) = 1 - \prod_{p \in \mathcal{P}} \bigl(1 - \text{comb}_p(n)\bigr),
\end{equation}
where $\mathcal{P}$ is the set of base primes. A final exponential decay
$s(n) = \exp(-\alpha \cdot \text{composite}(n))$ maps score 0 to 1 (prime) and 1 to 0
(composite). The decay parameter $\alpha$ controls sharpness and may be made learnable.
Detection is exact for all $n \leq N$ when $\max_p = \lfloor\sqrt{N}\rfloor$.

\subsection{Modular Membership via \texttt{radbas}}

A recurring primitive is soft equality testing via the radial-basis activation
$r(x) = \exp(-(\varepsilon x)^2)$, which peaks at $x{=}0$ \citep{paper1}.
For membership queries---``is $n$ a multiple of $p$?''---we precompute known multiples
as a lookup table, subtract the query, and apply \texttt{radbas}. The score is near 1
for exact matches and near 0 elsewhere, with no training required.

% ─────────────────────────────────────────────────────────────────────────────
\section{Experiments}
\label{sec:experiments}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Tier-1: Zero-Shot Structural Tasks}
\label{sec:tier1}

We evaluate three tasks for which the structured module achieves provable 100\% accuracy
without any training data. In each case we compare against an MLP baseline trained with
Adam \citep{adam2015} on the same task.

\paragraph{Task A --- Primality detection.}
Input: integer $n \in [2, 500]$.
Label: \texttt{is\_prime}$(n)$.
The structured module is \texttt{DifferentiableSieve(500)}.
The MLP baseline (3 hidden layers, 128 units, ReLU) is trained on $n \in [2, 100]$ for
300 epochs and evaluated on the full range.
Results in Table~\ref{tab:tier1} and Figure~\ref{fig:sieve} show that the sieve
achieves 100\% accuracy everywhere by construction; the MLP reaches 76.8\% in-range
and 82.5\% out-of-range---never reliably detecting primes, and requiring 2.2\,s of
training and 2{,}209 parameters to do so.

\paragraph{Task B --- Binomial expansion.}
Input: $(A, B, n)$ with $A, B \in \{-8,\ldots,8\}\setminus\{0\}$, $n \in \{2,\ldots,8\}$.
Label: coefficient vector of $(Ax + By)^n \in \mathbb{Z}^{16}$.
The dataset is a deterministic sweep of 1{,}792 samples (all valid triples), split
80/20 for the MLP baseline.
The structured module (\texttt{BinomialExpansion}) achieves 100\% exact accuracy on
all samples with zero training. The MLP (4 hidden layers, 256 units) achieves 0.3\%
exact accuracy on the validation split after 500 epochs: it approximates the
leading-coefficient magnitude but cannot recover integer polynomial coefficients
reliably. Figure~\ref{fig:binomial} shows the failure mode directly.

\begin{table}[t]
  \centering
  \caption{Tier-1 results. \emph{Exact} = fraction of outputs matching ground truth
    to integer precision. Sieve accuracy is provable; MLP accuracy is empirical.}
  \label{tab:tier1}
  \small
  \begin{tabular}{llrrrrr}
    \toprule
    Task & Method & Params & Train (s) & In-range & OOD & Exact (val) \\
    \midrule
    Primality   & \texttt{DifferentiableSieve} & \textbf{0} & \textbf{0} & \textbf{100.0\%} & \textbf{100.0\%} & --- \\
    Primality   & MLP baseline                 & 2{,}209    & 2.2        & 76.8\%            & 82.5\%            & --- \\
    \midrule
    Binomial    & \texttt{BinomialExpansion}   & \textbf{0} & \textbf{0} & ---               & ---               & \textbf{100.0\%} \\
    Binomial    & MLP baseline                 & 202{,}512  & 30.9       & ---               & ---               & 0.3\%  \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Tier-2: MoE Range Extrapolation}
\label{sec:tier2}

We train a mixture-of-experts primality model consisting of:
(i)~a \texttt{DifferentiableSieve} expert (0 learnable parameters; output detached
from the gradient graph);
(ii)~a blind MLP expert (2 hidden layers, 64 units);
(iii)~a learned router MLP (1 hidden layer, 32 units; softmax output $\mathbf{w}$).
The combined output is $\hat{y} = w_0 \hat{y}_\text{sieve} + w_1 \hat{y}_\text{blind}$.
All components are trained jointly on $n \in [2, 100]$ for 400 epochs with binary
cross-entropy loss. The sieve component receives no gradient.

The structured expert is exact on all $n \leq 500$ by construction; the blind MLP is
trained only on $[2, 100]$ and extrapolates approximately.
The central question is whether the router \emph{discovers} the asymmetry.

Figure~\ref{fig:router} (left) shows the mean router weight toward the sieve expert
$w_0$ over training. Starting from near-uniform initialisation ($w_0 \approx 0.58$),
the router weight rises monotonically, settling at $w_0 = 0.94$ by epoch 400.
This occurs without any supervision signal about which expert is structurally correct.
Figure~\ref{fig:router} (right) shows the per-$n$ router allocation at inference: the
router assigns high sieve weight uniformly across the input range, including the
out-of-distribution region $n > 100$.

Table~\ref{tab:tier2} summarises accuracy by range.
The sieve is exact everywhere; the MoE marginally exceeds the blind MLP in the
out-of-distribution range by inheriting the sieve's structural correctness.
The router does not fully collapse to $w_0 = 1$ because the sieve outputs soft scores
in $(0, 1)$ rather than hard binary labels---the 6\% residual reflects the MLP
providing a calibration hedge at the sigmoid boundary.

\begin{table}[t]
  \centering
  \caption{Tier-2 MoE range extrapolation. Training range: $n \in [2, 100]$;
    evaluation range: $n \in [2, 500]$.}
  \label{tab:tier2}
  \small
  \begin{tabular}{lrrrr}
    \toprule
    Method & Params (learned) & In-range [2,100] & OOD [101,500] & All [2,500] \\
    \midrule
    \texttt{DifferentiableSieve} & \textbf{0}   & \textbf{100.0\%} & \textbf{100.0\%} & \textbf{100.0\%} \\
    MoE (sieve + MLP + router)   & 4{,}483      & 77.8\%           & 82.5\%           & 81.4\%           \\
    Blind MLP                    & 4{,}353      & 77.8\%           & 82.5\%           & 81.4\%           \\
    \bottomrule
  \end{tabular}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section{Discussion}
\label{sec:discussion}
% ─────────────────────────────────────────────────────────────────────────────

\paragraph{Why not just compute it symbolically?}
The point is composability and differentiability. A frozen sieve buffer inside a neural
network serves as a feature extractor whose output flows into a learned classifier or
router. The structured module handles the exact part; the learned component handles
variation and uncertainty around it. This asymmetry mirrors how humans combine mental
arithmetic (fast, exact, bounded domain) with estimation or tool use (slower, flexible,
unbounded).

\paragraph{Relationship to the Peano axioms.}
Our modules demonstrate that combinatorial structures central to mathematics emerge from
Peano-like primitives: a single seed value and a fixed local rule, encoded as frozen
network weights. \texttt{PascalTriangle} starts from a single 1 and derives every
binomial coefficient by one step of the successor-like convolution
$\mathbf{K} = \bigl[\begin{smallmatrix}1&1\\0&0\end{smallmatrix}\bigr]$.
\texttt{BernoulliTriangle} applies a column-wise cumulative sum to the result.
\texttt{DifferentiableSieve} sweeps multiples of each prime in the same incremental
spirit. The underlying claim---that rich mathematical structure can be bootstrapped from
minimal local rules---is as old as Peano and as new as the observation that the same
operations are expressible as differentiable network layers.

\paragraph{Extension to logical inference.}
The \texttt{polyweave} library already ships a differentiable forward chainer over
propositional Horn knowledge bases (\texttt{polyweave.reasoning.ForwardChainer}), which
uses the product t-norm as a differentiable AND gate \citep{paper1}.
This module constitutes a natural third arm of the MoE: a \emph{logic expert} that
handles entailment queries exactly, alongside the arithmetic experts demonstrated here.
A router would then learn to distinguish numeric from logical queries---a step toward
the unified differentiable reasoning system sketched in Figure~\ref{fig:architecture}.
We leave this combination to future work, noting that all components are already
implemented and publicly available.

\paragraph{Toward a Large Maths Model.}
The MoE pattern demonstrated here scales to a richer expert pool: modular arithmetic
(\texttt{radbas} membership), recurrence relations (fixed-weight RNN), sorting
(differentiable sorting networks), and GCD (bounded Euclidean unrolling) are all
expressible as parameter-free modules in the same framework. Combined with a
natural-language front-end---analogous to the convolutional LaTeX parser prototyped
alongside this work---such a system would constitute a \emph{Large Maths Model}: an
architecture in which a language model delegates well-posed arithmetic queries to
exact internal modules rather than approximating them in its feed-forward layers.

\paragraph{Limitations.}
The modules are exact only within their specified ranges ($N$ for the sieve;
\texttt{num\_rows} for Pascal). The MoE router in Section~\ref{sec:tier2} uses
a single structured expert; multi-expert routing with task-type disambiguation is
future work. The router weight $w_0 = 0.94$ falls short of 1 because the sieve
outputs soft scores; a hard-threshold variant would achieve perfect routing accuracy
at the cost of non-differentiability.

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}
\label{sec:conclusion}
% ─────────────────────────────────────────────────────────────────────────────

We have shown that classical combinatorial algorithms are expressible as zero-parameter
neural network modules, enabling exact computation within differentiable architectures.
A learned router, trained only on classification loss, discovers without supervision
that it should delegate to the structured expert---a concrete demonstration that
structured and learned computation are complementary rather than competing.
The three modules presented here, together with the existing \texttt{polyweave.reasoning}
forward chainer, form the nucleus of a broader differentiable reasoning library
whose capabilities extend from arithmetic through to propositional logic.
We release all code and experiments at \url{https://github.com/swhipp/autodidact-hypernet}.

% ─────────────────────────────────────────────────────────────────────────────
\section*{Acknowledgements}

The author thanks the open-source communities behind PyTorch, NumPy, and Matplotlib.

% ─────────────────────────────────────────────────────────────────────────────
\bibliographystyle{plainnat}
\bibliography{references_paper3}

% ─────────────────────────────────────────────────────────────────────────────
\appendix

\section{Module Implementation Details}
\label{app:impl}

All modules are available as \texttt{polyweave.maths.\{PascalTriangle, BinomialExpansion,
BernoulliTriangle, DifferentiableSieve\}}. Each inherits from \texttt{nn.Module} and
registers its structural weights as frozen buffers via \texttt{register\_buffer}.
The \texttt{forward()} method is decorated with \texttt{@torch.no\_grad()} where no
downstream gradient is required.

\paragraph{PascalTriangle.} $O(N^2)$ memory; $O(N)$ forward passes of a single
$1{\times}1{\times}2{\times}2$ convolution. Exact for $n \leq 15$ in \texttt{float32}.

\paragraph{DifferentiableSieve.} $O(N \log \log N)$ buffer storage (by the prime
harmonic series). Decay parameter $\alpha$ defaults to 5.0; values $\geq 10$ give
near-binary output.

\paragraph{BernoulliTriangle.} Single \texttt{torch.cumsum} call on the Pascal output;
negligible additional cost.

\section{Extended Results}
\label{app:results}

\begin{table}[h]
  \centering
  \caption{Tier-1 primality: per-range breakdown.}
  \small
  \begin{tabular}{lrrrr}
    \toprule
    Method & [2, 50] & [51, 100] & [101, 250] & [251, 500] \\
    \midrule
    \texttt{DifferentiableSieve} & 100.0\% & 100.0\% & 100.0\% & 100.0\% \\
    MLP baseline & [TBD] & [TBD] & [TBD] & [TBD] \\
    \bottomrule
  \end{tabular}
\end{table}

\end{document}
